### 2020-09-15 submission/manuscript.tex
First pass through Results

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\vspace*{10mm}

\hypertarget{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}{%
\section{Removal of rare sequences from 16S rRNA gene sequence surveys
biases the interpretation of community structure
data}\label{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}}

@@

Patrick D. Schloss\({^1}\)\({^\dagger}\)

@@

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

\(1\) Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI 48109

@@

\hypertarget{observation-format}{%
\subsection{Observation format}\label{observation-format}}

@@

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

@@

\hypertarget{importance}{%
\subsection{Importance}\label{importance}}

@@

Question: What impact does removing singletons (and other rare
sequences) have on downstream data analysis?

\begin{itemize}
\tightlist
\item
  Many pipelines remove rare sequences prior to clustering sequences
  into OTUs or as part of the generation fo ASVs
\item
  Singletons can be sequences that appear 1-10 times in larger samples
\item
  Seeks to solve the problem of low quality sequence data - more
  singletons with higher error rates
\item
  Likely there are rare sequences that are good - e.g.~high diversity
  samples
\item
  People have overfit their pipelines ot mock community data which have
  unrealistic community distributions (i.e.~uniform abundance or log
  abundance) and phylogenetic diversity
\item
  Better solutions would include:

  \begin{itemize}
  \tightlist
  \item
    Get better sequence data
  \item
    Rarefy data so that the number of errors are controlled for across
    samples
  \end{itemize}
\item
  Not all samples are sequenced evenly, so a singleton doesn't hav the
  same relative abundance across all samples
\end{itemize}

H: Removing rare sequences from each sample will skew the community
distribution to impact the amount of variation in the relative abundance
and alpha and beta diversity values across samples

@@

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

@@

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

@@

\hypertarget{references}{%
\subsection{References}\label{references}}

\hypertarget{refs}{}

@@

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\subsection{word choice: sample
vs.~community}\label{word-choice-sample-vs.community}

\section{Removal of rare sequences from 16S rRNA gene sequence surveys
biases the interpretation of community structure
data}\label{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}

@@

Patrick D. Schloss${^\dagger}$

@@

$\dagger$ To whom correspondence should be addressed:

\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

Department of Microbiology and Immunology University of Michigan Ann
Arbor, MI 48109

@@

\subsection{Research article format}\label{research-article-format}

@@

\subsection{Abstract}\label{abstract}

250 words

@@

\subsection{Importance}\label{importance}

150 words

@@

\textbf{EDIT: Two elements that are held in tension in the analysis of
16S rRNA gene sequence data is how to adequately remove PCR and
sequencing artifacts and decrease the granularity of the taxonomic level
that is used in the analysis.} Previous attempts have included screening
for sequencing quality based on quality scores {[}Kozich/Edgar{]}
followed by a polishing step based on the frequency of the sequences
relative to similar sequences {[}Kozich/Edgar/DeBlur{]}. Other pipelines
model the quality scores and types of errors to cluster sequences
directly {[}Dada2{]}. But, as a final step many pipelines advocate for
removing rare sequences from each dataset prior to outputting the
sequence data as amplicon sequence variants (ASVs)
{[}Knight/Edgar/DeBlur/Dada2{]}. ASVs are often clustered further to
generate operational taxonomic units or phylotypes. Some pipelines
remove all ASVs that appear once (i.e.~singletons) {[}XXXX{]}, XXXXXX
{[}XXXXXX{]}, XXXXX {[}XXXXXXX{]}, or XXXXXX {[}XXXXXX{]} times prior to
further clustering or making ecological comparisions. Notably, the
mothur-based pipeline discourages the practice of removing rare
sequences.

The abundance-based screening approach assumes that rare ASVs are more
likely to be artifacts than more abundant ASVs. Sequencing of mock
communities confirms that artifacts tend to be rare. Proponents of
abundance-based screening point to their ability to obtain the correct
number of ASVs, OTUs, or phylogypes with data generated from sequencing
mock communities when rare ASVs are removed. However, this approach
effectively overfits the curation pipeline to data generated from a
phylogenetically simple community with an atypical community
distribution that is often sequenced to a depth that is not achieved
with biological samples. It is necessary to think more deeply about the
practice of abundance-based screening.

The minimum abundance thresholds that have been proscribed are developed
and applied without regard for the total number of sequences generated
from each sample. Ignored in their recommendations is the common
experience that the number of sequences generated from each sample may
vary by two or three orders of magnitude. An ASV that appears once in a
sample with 2,000 sequences is more trustworthy than an ASV that appears
once in a sample with 100,000 sequences since it has a 50-fold higher
relative abundance. But, according to the pipeline recommendations, they
are treated as being equally trustworthy. Rather than removing rare
ASVs, the approach taken by the mothur pipeline applies the classical
ecological approach of rarefaction. Each sample is rarefied to the same
sequencing depth so that the number of artifacts that appears in each
sample is controlled.

Experience sequencing biological samples demonstrates that there are
good ASVs that may have an abundance below the proscribed threshold. For
example, the abundance of an ASV may be below the threshold in some
samples or time points and above the threshold in others. However,
rarity, both in terms of prevalence and incidence, is an important
ecological concept. Removing rare ASVs likely hinders one's ability to
ability to make inferences about the dynamics and nature of the
populations that rare ASVs represent. Furthermore, removing ASVs whose
abundances are below the proscribed threshold also potentially biases
the community structure of the samples.

In the current study, I use published sequence data from 12 studies to
investigate the nature of rare ASVs (i.e.~those that appear 10 or fewer
times) and the effects of removing them on downstream analysis of
microbial communities. The analysis was also performed using traditional
operational taxonomic units, where ASVs subjected to abundance-based
screening were clustered such that the ASVs within an OTU were no more
than 3\% different from each other. The results reject the assumptions
built into abundance-based screening and highlight the problems inherent
in removing rare ASVs.

\textbf{\emph{Results.}}

\textbf{Datasets.} I collected 12 publicly available datasets that used
the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA
gene from a variety of environments (Table 1). To insure the highest
possible data quality, datasets were limited to those where the 500
cycle v2 MiSeq chemistry was used to sequence the amplicons. The paired
250 nt reads resulted in near complete 2-fold sequencing coverage of
every nucleotide in the ca. 250 nt-long region. This region and
sequencing platform were selected because previous work has shown that a
standard data analysis pipeline in mothur results in a sequencing error
rate below 0.02\% (\textbf{???}). All sequence data were obtained from
the Sequence Read Archive and processed using a standard mothur-based
sequencing pipeline that resulted in ASVs as generated by the
pre.cluster algorithm (\textbf{???}, 1). After removing poor quality and
chimeric ASVs and samples that had uncharacteristically low number of
sequences for the dataset, these datasets included between 7.00 and
490.00 samples. The median number of sequences for each dataset ranged
between 6,477.00 and 193,464.00 (Table 1, Figure S1). Strikingly, aside
from the relatively small marine and soil datasets, the difference
between the sample with the fewest sequences and the sample with the
most sequences for each dataset varied between 7.4 and 96.6-fold (Table
1).

\textbf{The nature of singletons.} Removal of rare ASVs is commonly
justified as a method of removing ASVs that are artifacts. If such ASVs
are artifacts, then one would expect the number of singleton ASVs to
accumulate with sequencing depth. Contrary to this expectation, the
median percentage of sequences that were discarded when singleton ASVs
were removed from each dataset varied between 0.42 and 22.23\%
(bioethanol and seagrass). In addition, with the exception of the
samples from the marine and sediment datasets (Spearman correlation,
P\textgreater{}0.05), the fraction of singleton ASVs in the sample was
negatively correlated with the number of sequences in each sample with a
range between -0.27 and -0.87 (rice and bioethanol). This showed that
with additional sequencing, the probability of seeing rare ASVs multiple
times was greater than the probability of generating an artifact. This
suggests that the rare ASVs are not as likely to be artifacts as
previously thought. Furthermore, if rare ASVs were artifacts, then one
would not expect to find them in other samples from the same dataset. In
fact, singleton ASVs from samples with fewer sequences were often found
in samples with more sequences. At least 50\% of the singleton ASVs
found in the samples from the mice, rice, seagrass, and stream datasets
were found in another sample from the same dataset. Considering the
likelihood of finding an ASV duplicated in another sample is confounded
by the number of samples and inter-sample diversity, the high coverage
of singleton ASVs in these datasets was remarkable. The correlation
between the number of sequences in a sample and the fraction of that
sample's singleton ASVs that were covered by another sample in the
dataset was significant and negative for 9 of the datasets ranging
between -0.31 and -0.84 for the rice and seagrass datasets,
respectively. The negative correlation indicated that the singleton ASVs
in the smaller samples were more likely to be covered by ASVs in the
larger samples. Among the three datasets without a significant
correlation (Spearman correlation, P\textgreater{}0.05), the marine and
soil datasets had the fewest samples in our collection and the stream
dataset already had a high level of coverage regardless of the number of
sequences. Contrary to the common motivation for removing rare ASVs,
these results indicate that this practice disproportionately impacts
samples with fewer sequences and likely removes a large number of
non-artifact ASVs.

\textbf{How do communities change when rare ASVs are removed?} Removing
rare ASVs will reduce the number of observed ASVs and proportionally
increase the relative abundance of the remaining ASVs. The result was
expected to be a loss of information contained within each sample. To
quantify the effect of removing rare ASVs on the information contained
within each sample, I varied the minimum abundance threshold to simulate
removing ASVs of varying rarity from each sample. The number of ASVs in
each sample decreased by between 34.4 and 86.2\% when removing those
ASVs that only appeared once and by between 76.0 and 95.6\% when
removing those that appeared ten or fewer times from each sample (Figure
1). Similarly, the Shannon diversity decreased by between 1.8 and 15.9\%
when removing ASVs that only appeared once and by between 76.0 and
95.6\% when removing ASVs that appeared ten or fewer times from each
sample (Figure 1). Next, I assigned the ASVs to OTUs to assess the
impact of removing rare ASVs on higher level taxonomic groupings that
are commonly used in microbial ecology studies. Although pooling similar
ASVs into OTUs reduced the impact of removing the rare ASVs relative to
the ASV-based analysis, the minimum abundance threshold still decreased
the number of observed OTUs and the diversity decreased relative to the
original community (\textbf{Figure 1}). In contrast to the number of
ASVs or OTUs and their diversity, the Kullback--Leibler divergence
compares the relative abundance of each ASV or OTU between
representations of the community. I calculated the Kullback--Leibler
divergence from the full to pruned communities when rare ASVs were
removed. As the threshold for removing ASVs increased, the amount of
information lost also increased for both ASVs and OTUs (Figure 1). The
relative loss of information was generally lower for OTUs than than it
was for ASVs. Removing rare ASVs, regardless of abundance threshold, had
profound impacts on the representation of the communities.

\textbf{Removing treatment group effects from community data.} Because
treatment effects often affect a sample's diversity and inter-sample
variation, I generated null distributions for each study by randomizing
the number of times each ASV was observed in each sample such that the
total number of sequences in each sample and the total number of times
each ASV was observed across all samples in the study was the same as
was originally observed. This effectively made every community in a
study a statistical sample of the study-wide composite community
distribution. For example, after this procedure, the 490 samples from
the human dataset would be expected to have the same number of ASVs and
diversity and one would not expect to find treatment-based effects
between the samples. Because of the risk of bias if only one
representation of the null distribution was generated, I generated 100
randomized datasets for each study. The same trends between removing
rare ASVs and the number of ASVs, sample diversity, and information loss
that were identified using the observed community community distribution
data were also identified with the data from the null distribution
(\textbf{Figure S2}). The null distribution data were used in the
remainder of the study.

\textbf{How do the relationships between communities change when they
are screened by removing rare ASVs.} Considering the loss of the number
of ASVs, diversity, and information when a community has its rarest ASVs
removed, it seemed likely that the relationship between communities
would be altered. To assess the impact of removing rare ASVs on measures
of alpha diversity between samples I calculated the coefficients of
variation (COV, i.e.~standard deviation divided by the mean) for the
number of ASVs and diversity for each study at multiple abundance
thresholds. The COV for the number of ASVs across the studies after
removing singletons were between 3.6 and 32.7-times larger than they
were without removing singleton ASVs. Similarly, the COVs for the
diversity of ASVs were between 1.8 and 20.4-times larger when singletons
were removed than when they were not removed.To assess the impact of
removing rare ASVs on measures of beta diversity between samples I
calculated the COV of the Bray-Curtis distances between samples within
the same study at multiple abundance thresholds. The COV between
Bray-Curtis distances within a study when singletons were removed was
between 1.3 and 18.6-times larger than when they were not removed.
Increasing the minimum abundance threshold increased the COV between
samples when using metrics of alpha and beta diversity. When ASVs were
clustered into OTUs the difference in the COV was less than it was for
the ASVs (\textbf{Figure 2}). These results indicate that removing rare
ASVs increases the dissimilarity between samples, which could have a
significant impact on the statistical power to detect differences
between treatment groups.

\textbf{How does the increased inter-sample variation impact the ability
to detect differences between samples?} To test the effect of increased
inter-sample variation, I randomly assigned samples to one of two
treatment groups. In the first treatment group, communities were
randomly sampled from the null distribution as described above. For the
second treatment group, I increased the abundance of 10\% of the ASVs in
the pooled study distribution by 5\%. I randomly generated 100 simulated
sets of treatment groups and samples. I then tested the ability to
detect a difference between the two treatment groups using alpha and
beta diversity metrics. The fraction of significant tests was a
measurement of the statistical power to detect the difference between
the treatment groups. Across all of the metrics I tested, the marine
dataset yielded no simulated sets that were statistically significant,
which was likely due to the small number of samples in the study (N=7);
the marine dataset is not included in the following analysis. Among the
remaining datasets, the power to detect a difference in the observed
number of ASVs ranged between 0.10 and 0.49 and between 0.10 and 0.53 to
detect a difference in diversity when using a Wilcox test
(\textbf{Figure X}). When singleton ASVs were removed, the power to
detect a difference in the number of ASVs dropped by between 27.3 and
92.9\% and by between 40.0 and 93.3\% to detect a difference in
diversity. The effect of removing rare ASVs on the number of OTUs and
their diversity was similar. A popular alternative approach calculates
the dissimilarity between communities based on the abundance of each ASV
or OTU. We used the Bray-Curtis dissimilarity index to compare the
simulated communities within each dataset and calculated the power to
detect differences between the two simulated treatment groups using the
analysis of molecular variance (also called PERMANOVA). Without removing
rare sequences, the power to detect a difference between the two
simulated treatment groups varied between 0.41 and 1.00. Aside from 3
datasets, the power to detect differences dropped by between 6.5 and
64.0\% when singletons were removed. However, when ASVs that occurred 10
or fewer times were removed from each sample, the power to detect
differences dropped by 12.0 and 97.2\%; similar results were observed
when ASVs were clustered into OTUs. Removing rare ASVs reduced the
ability to detect simulated treatment effects using metrics commonly
used to compare microbial communities.

\textbf{How does the removing of rare ASVs impact the probability of
falsely detecting a difference between treatment groups?} Observing
reduced ability to detect differences between communities when rare ASVs
were removed from each sample, I next asked whether removing rare ASVs
could lead to falsely claiming that a treatment effect had a significant
effect on community diversity and structure. First, for each dataset, I
sampled sequences from the null distribution and randomly assigned each
sample to one of two treatment groups and determined the number of
observed ASVs and OTUs and their Shannon diversity index. Testing at an
experiment-wise error rate of 0.05, I expected 5\% of the iterations for
each dataset to yield a significant test result. Indeed, there was no
evidence that removing rare ASVs resulted in an inflated experiment-wise
error rate. The average fraction of significant tests did not
meaningfully vary from 0.05 across the minimum abundance threshold,
dataset, metric of describing sample alpha-diversity, or whether the
abundance of ASVs or OTUs were used (\textbf{Figure XA}). Similarly, the
average fraction of significant tests did not meaningfully vary from
0.05 when using analysis of molecular variance to compare communities
using Bray-Curtis distances. Second, I again sampled sequences from the
null distribution, but assigned samples to one of two treatment groups
based on the number of sequences in each sample. The samples with fewer
than the median number of sequences for the dataset were assigned to one
group and those with more than the median were assigned to the other.
This exaggerated bias has been observed in comparisons of the lung and
oral microbiota because of the larger nubmer of non-specific amplicons
that can be sequenced from lung samples relative to those in the oral
cavity {[}\textbf{REF}{]}. When rare sequences were not removed, the
fraction of significant tests did not differ from 5\% for comparing the
number of observed taxa (using ASVs or OTUs), their shannon diversity,
or Bray-Curtis distances. However, when rare taxa of any frequency were
removed, the probability of falsely detecing a difference as signifiant
increased with the definition of rarity (\textbf{Figure XB}). Across the
datasets, the average fraction the average fraction of falsely detecting
a difference was when only singletons were removed was greater than
90\%. If there is any relationship between the number of sequences and
the treatment group, the risk of falsely rejecting the null hypothesis
is inflated when researchers use the strategy of removing rare
sequences. The most conservative approach is to not remove low abundance
sequences.

\textbf{\emph{Conclusion.}} Removing rare sequences decreases the
diversity represented by 16S rRNA gene sequence data and increases the
variation between samples. Such impacts will hinder the statistical
power to differentiate between treatment groups. Instead of removing
rare sequences, researchers should focus on optimizing their sequence
generation to minimize the amount of PCR and sequencing errors. In
addition, samples should be rarefied to a common number of sequences
across samples without prior culling of rare sequences. The number of
artifacts is correlated to the number of sequences being considered.
With this in mind, rarefaction allows one to control for uneven sampling
effort and to control for the number of artifacts in the analysis.

Need to treat rare sequences with a grain of salt

@@

\subsection{Acknowledgements}\label{acknowledgements}

@@

\subsection{Materials and Methods}\label{materials-and-methods}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  sequencing pipeline description
\end{itemize}

@@

\subsection{References}\label{references}

@@

1. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{http://dx.doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
